---
title: "R Workshop"
author: "Brett Johnson"
date: "1/21/2019"
output: html_document
---

## Creating objects in R

```{r setup, warning=FALSE}
library(tidyverse)
#read in data files and load libraries in setup chunk
```


```{r arithmetic}

3 + 5

12 / 7 

5 * 5

# assign value to object 
weight_kg <- 55

2.2 * weight_kg

weight_lb <- 2.2 * weight_kg

sqrt(weight_kg)

round(3.14159)

round(3.141159, digits = 2) 

round(3.14159, 2)
```

### Vectors and data types

This section will describe some basic data types in R

```{r}

weight_g <- c(50,60,65,82)

animals <- c("mouse", "rat", "dog")
```

Vector types in R:
* numeric
* character
* logical (TRUE or FALSE)
* Factors (categorical data) 
* Dates 

A vector is a data structure in R.

Other data structures:
* lists
* data frames
* matrices
* of course vectors

Often you want to conver lists and matrices to data frames or vectors.

## Data Frames

Next we're going to look at the structure of Data Frames.

```{r}
library(tidyverse)

download.file(url="https://ndownloader.figshare.com/files/2292169", destfile = "read_data/portal_data_joined.csv") 

library(here) # This package makes working directories and file paths easy

surveys <- read_csv(here("read_data", "portal_data_joined.csv"))

surveys

str(surveys)

dim(surveys)
nrow(surveys)
ncol(surveys)

summary(surveys)
```

### Indexing and subsetting data frames

First lets use square bracket subsetting.

Square brackets are great for defining coordinates to extract data from. But what happens when the structure of the data frame. 

```{r}

# First define the row coordinate, and then the column. Also write row and then column
surveys[1,1]

surveys[1,6]

# Defining only which element we want will return a data frame
surveys[1]

surveys[1:3, 7]

# Give us all the rows and columns except column 7
surveys[,-c(1:5)]
```

## Data Manipulation

Key functions for data manipulation:

* select(): subsetting columns
* filter(): subsets of rows based on conditions
* mutate(): create new columns, based on information from other columns 
* group_by(): creates groups based on categorical data in a column 
* summarize(): creates summary stats on grouped data
* arrange(): sort results
* count(): gives a count of discrete values

```{r}
select(surveys, plot_id, species_id, weight)

# Negative subsetting
select(surveys, -record_id)

filter(surveys, year == 1995)


```

## Pipes

Pipes allow you to chain together dplyr functions.

Pipe: %>% or ctrl-shift-m

```{r}
# Write multiple arguments in a sentence using pipes
surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)

surveys_sml <- surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)
```


Challenge # 1

Using pipes, subset the surveys dataframe to include animals collected 1995 an retain only the columns year, sex and weight.

```{r challenge 1}

```

```{r}
surveys %>% 
  mutate(weight_kg = weight / 1000,
         weight_kg2 = weight_kg * 2)

surveys <- surveys %>% 
  drop_na(weight) %>% 
  mutate(mean_weight = mean(weight))

mean(surveys$weight)
```

Challenge # 2

Using the surveys data frame create a new dataframe contains only the species_id column, has a new column called hindfoot_half: contains values that are half the hindfoot_length values. Also, in the new hindfoot_half column there are no NAs and values are all less than 30.

```{r}
surveys_hindfoot_half <- surveys %>% 
  drop_na(hindfoot_length) %>% 
  mutate(hindfoot_half = hindfoot_length / 2) %>% 
  filter(hindfoot_half < 30) %>% 
  select(species_id, hindfoot_half)
```


```{r}
surveys %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE)) 

# arrange by defaualt is done in ascending order
surveys %>% 
  group_by(sex, species_id) %>% 
  summarize(mean_weight = mean(weight, na.rm = TRUE),
            min_weight = min(weight, na.rm = TRUE)) %>% 
  arrange(min_weight)

```


```{r}
surveys %>% 
  count(sex, sort = TRUE)

# the above code is synonomous
surveys %>% 
  group_by(sex) %>% 
  summarize(count = n())

```

Challenge # 3

1. How many animals were caught in each plot_type surveyed.

2. Use group_by and summarize to find the mean, min, and max of hindfoot length (using species_id) for each species. Also, add the number of observations (hint: see ?n)

```{r}
surveys %>% 
  group_by(species_id) %>% 
  summarize(mean_length = mean(hindfoot_length, na.rm = TRUE),
            n())
  
```

3. What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and weight.

```{r}

surveys %>% 
  group_by(year, genus, species_id) %>% 
  summarize(max_weight = max(weight, na.rm = TRUE)) %>% 
  arrange(desc(max_weight)) 

max_weights <- surveys %>% 
  drop_na(weight) %>% 
  group_by(year) %>% 
  filter(weight == max(weight)) %>% 
  select(year, genus, species_id, species, weight) %>% 
  arrange(year) %>% 
  unique()

```

## Export our data
```{r}
write_csv(max_weights, here("write_data", "max_weights.csv"))
```

# Day 2

## Tidy Data in Spreadsheets

Today we are going to look at tidying data.

The functions we use for tidying data are:

* tidyr::spread()
* tidyr::gather()

### spread()

spread() takes three principal arguments:
1. the data
2. the _key_ column variable will become the new column names
3. the _value_ column variable which will fill the new column variables 

Were going to use the surveys dataset

```{r}
library(tidyverse)
surveys <- read_csv(here::here("read_data", "surveys.csv"))

# create a wide data format of surveys using spread

surveys_gw <- surveys %>%
  drop_na(weight) %>% 
  group_by(species_id) %>% 
  summarize(mean_weight = mean(weight))

str(surveys_gw)

# Now we want to spread the data set out into a wide format

wide_surveys_gw <- surveys_gw %>% 
  spread(key = species_id, value = mean_weight)

# Now we are going back to long from the wide data set that we just created.
```

### gather()

gather() takes four arguments:
1. data
2. key
3. value
4. names of columns we use to fill the key variable (or drop)

```{r}
long_surveys <- wide_surveys_gw %>% 
  gather(key = species_id, value = mean_weight) 
```

## Sending Tidy Data

### Changelog

* Update your changelog with changes you make to your raw data, or other significant changes or additions to your projects!

### Data Dictionary
* Create a data dictionary to define our variables


```{r}
tidy_gsi <- read_csv(here::here("write_data", "tidy_gsi.csv"))
```


# Analyzing Data

## Importing from the Hakai Data Portal

Switched to data_wrangling script to import data into our read_data file.

We have chla data, fish data, and sockeye stock ID data. Let's start to analyze these data sets.

```{r}
fish <- read_csv(here::here("read_data", "fish.csv"))
chla <- read_csv(here::here("read_data", "chla.csv"))
tidy_gsi <- read_csv(here::here("write_data", "tidy_gsi.csv"))
```

```{r}
fish %>% 
  count(species)

fish_d09 <- fish %>% 
  filter(site_id == "D09") %>% 
  select(hakai_id, jsp_survey_id, seine_id, date, species, site_id, fork_length, weight) %>% 
  mutate(k = (10 ^ 5 * weight) / fork_length ^ 3) %>% 
  drop_na(k)
```

## Factors

```{r}
str(fish_d09)

class(fish_d09$species)

# coerce a column to be a factor, do this:

fish_d09$species <- factor(fish_d09$species)

class(fish_d09$species)

levels(fish_d09$species)
```

If you have factors that are numbers, don't try to do math with those factors. Under the hood R will be treating your factor levels as a different number.

## Dates

read_csv() treats ISO date standards (yyyy-mm-dd) as a DATE object
read.csv() treats them as characters: not ideal

Lubridate is a package for dealing with dates.

```{r}
library(lubridate)

fish_d09$date <- as_date(fish_d09$date)

class(fish_d09$date)


# extract date components
fish_d09 <- fish_d09 %>% 
  dplyr::mutate(year = year(date),
         month = month(date),
         week = week(date),
         yday = yday(date))


```

Lubrudate you can do math with dates

* periods
* intervals
* durations

# Joining Data

Dat we have:

* chla
* tidy_gsi
* fish_d09

```{r}
left_join(fish_d09, tidy_gsi, by = 'hakai_id') 

right_join(fish_d09, tidy_gsi) 

so_gsi <- inner_join(fish_d09, tidy_gsi, by = 'hakai_id') 

anti_join(fish_d09, tidy_gsi)
```

## ggplot2

To build a ggplot:

ggplot(data = DATA, mapping = aes(MAPPINGS))
  + GEOM_FUNCTION()

```{r, eval = FALSE}
# EXAMPLE
 ggplot(data = surveys, mapping = aes(species, weight)) +
  geom_point() 
```

```{r}
ggplot(so_gsi, aes()) +
  geom_point(aes(x = yday, y = fork_length)) +
  geom_smooth(aes(x = yday, y = fork_length), method = lm) +
  theme_classic() + 
  facet_wrap(. ~ stock_1)
  
```

```{r}
fish %>% 
  ggplot(aes(x = species, y = fork_length)) +
  geom_boxplot() +
  facet_wrap(year(date) ~ .)
```

```{r}
chla %>% 
  filter(work_area == "QUADRA") %>% 
  ggplot(aes(x = date, y = log(chla_final))) +
  geom_point(alpha = 0.5) +
  geom_smooth()
```


Cookbook for R
How to adjust legends, make facets, adjsut axes

http://cookbook-r.com/Graphs

Geom list:

https://ggplot2.tidyverse.org/reference/